The Decline of US Birth Rates: Analyzing Minimum Wages, Abortion Policies, Education, and Child Support Enforcement

### 2024-02-15 paper/paper.qmd
submission

```{r echo=FALSE, message=FALSE, results='asis'}
#| label: tbl-1
# Define file paths 
policy_data_path <- "../data/analysis_data/annual_policy/policyvars01_19.dta"
population_data_path <- "../data/UKCPR_National_Welfare_Data_Update_020623.xlsx"
births_data_path <- "../data/analysis_data/annual_policy/numbirths_2001_2019.csv"

@@

@@

TODO: Results description for Education

@@

## Second discussion: Political affiliation and birth rates

@@

@@

(ACA)’s Medicaid is “intended to address systemic health inequalities
for millions of Americans who lack health insurance” Medicaid is
“expected to provide coverage to low-income individuals,” and also
covers low-income pregnant women. Excluding the states where abortion is
banned, Medicaid covers abortions, making it more accessible. According
to Figure C1, More people continue to have medicaid as seen based on the
trend from 2000 to 2020, making an affordable abortion more accessible.
In Figure C3, we can see that there is an increase in parental consent
for abortion. In other words, more parents are increasingly interested
in getting an abortion every year. A potential motivator for parents to
want an abortion is that as a low-income family, it is harder to take
care of a larger family. Welfare is “federal government-sponsored
assistance programs for individuals and families in need.” In Figure C2,
it is evident that there is a decrease in maximum welfare benefits,
making it less affordable for lower-income families to have a larger
family size.The increasing demand for an abortion and the insurance
covering it may contribute to a decline in birthrates.

@@

# References
Grundy E, Murphy M. Demography and public health. In: Detels R, Gulliford M, Karim QA, Tan CC, editors. Oxford textbook of global public health. 6th ed. Oxford, England: Oxford University Press, 718–35. 2015.

>>>>

```{r echo=FALSE, message=FALSE, results='asis'}
#| label: fig-11
# Define file paths 
policy_data_path <- "../data/analysis_data/annual_policy/policyvars01_19.dta"
population_data_path <- "../data/UKCPR_National_Welfare_Data_Update_020623.xlsx"
births_data_path <- "../data/analysis_data/annual_policy/numbirths_2001_2019.csv"

@@

# Plotting for the DiD analysis
ggplot(did_data, aes(x = year, y = relative_births, color = stname)) +
  geom_line() +
  geom_vline(xintercept = 2011, linetype="dashed", color = "red") + # Policy change year
  labs(title = "Relative Births in NC (Treatment) and TN (Comparison) Over Time",
       x = "Year", y = "Relative Births per 1000 People") +
  theme_minimal() +
                       theme(
                         plot.title = element_text(size = 10)
                        )

@@

The figure 6 displays the trend in number of births segmented by education achievement from 1990 to 2019. The education levels all follow a similar trend with the exception of less than HS. For this education level, there is a decline in number of births post 2008 whereas the other education levels showed an increase in birth rates.

@@

## Political affiliation and birth rates

@@

## Education
We observe a convergence in birth rates across different educational categories based on the data for number of births by education level. This could be a reflection of larger societal shifts in family planning preferences and educational achievement. College graduates' stable fertility rate followed by a fall in births could mean that this group is delaying or reducing the size of their family, possibly for financial or career-related reasons. The overall decline in births over the past few years across all educational levels may reflect a general trend toward smaller families, or it may be the result of other outside factors not shown in the graph, such as shifting social norms, shifting economic conditions, or expanding access to reproductive healthcare.

@@

(ACA)’s Medicaid is “intended to address systemic health inequalities
for millions of Americans who lack health insurance” Medicaid is
“expected to provide coverage to low-income individuals,” and also
covers low-income pregnant women. Excluding the states where abortion is
banned, Medicaid covers abortions, making it more accessible. According
to figure 8, More people continue to have medicaid as seen based on the
trend from 2000 to 2020, making an affordable abortion more accessible.
In figure 10, we can see that there is an increase in parental consent
for abortion. In other words, more parents are increasingly interested
in getting an abortion every year. A potential motivator for parents to
want an abortion is that as a low-income family, it is harder to take
care of a larger family. Welfare is “federal government-sponsored
assistance programs for individuals and families in need.” In figure 9,
it is evident that there is a decrease in maximum welfare benefits,
making it less affordable for lower-income families to have a larger
family size.The increasing demand for an abortion and the insurance
covering it may contribute to a decline in birthrates.

@@

# References
@Dataset{ukcprWelfareData,
  title = {UKCPR National Welfare Data, 1980-2021},
  author = {{University of Kentucky Center for Poverty Research}},
  year = {2023},
  month = {Feb},
  address = {Lexington, KY},
  url = {http://ukcpr.org/resources/national-welfare-data%7D,
  note = {Accessed: <Feb 15, 2024>},
}

@Manual{citeR,
    title = {R: A Language and Environment for Statistical Computing},
    author = {{R Core Team}},
    organization = {R Foundation for Statistical Computing},
    address = {Vienna, Austria},
    year = {2023},
    url = {https://www.R-project.org/},
  }

@InBook{Grundy2015Demography,
  author    = {Emily Grundy and Michael Murphy},
  editor    = {Roger Detels and Martin Gulliford and Quarraisha Abdool Karim and Chorh Chuan Tan},
  title     = {Demography and public health},
  booktitle = {Oxford Textbook of Global Public Health},
  publisher = {Oxford University Press},
  address   = {Oxford, England},
  edition   = {6},
  pages     = {718--735},
  year      = {2015},
}

@Article{Kearney2022Puzzle,
  author    = {Melissa S. Kearney and Phillip B. Levine and Luke Pardue},
  title     = {The Puzzle of Falling US Birth Rates since the Great Recession},
  journal   = {Journal of Economic Perspectives},
  year      = {2022},
  volume    = {36},
  number    = {1},
  pages     = {151--176},
  doi       = {10.1257/jep.36.1.151},
}

@Article{thereferencecanbewhatever,
    title = {Welcome to the {tidyverse}},
    author = {Hadley Wickham and Mara Averick and Jennifer Bryan and Winston Chang and Lucy D'Agostino McGowan and Romain François and Garrett Grolemund and Alex Hayes and Lionel Henry and Jim Hester and Max Kuhn and Thomas Lin Pedersen and Evan Miller and Stephan Milton Bache and Kirill Müller and Jeroen Ooms and David Robinson and Dana Paige Seidel and Vitalie Spinu and Kohske Takahashi and Davis Vaughan and Claus Wilke and Kara Woo and Hiroaki Yutani},
    year = {2019},
    journal = {Journal of Open Source Software},
    volume = {4},
    number = {43},
    pages = {1686},
    doi = {10.21105/joss.01686},
  }

  @Article{Nargund2009DecliningBirthRate,
    title = {Declining birth rate in Developed Countries: A radical policy re-think is required},
    author = {Nargund, G.},
    year = {2009},
    journal = {Facts Views Vis Obgyn},
    volume = {1},
    number = {3},
    pages = {191-193},
    PMID = {25489464},
    PMCID = {PMC4255510},
}

@Manual{citeLahman,
    title = {Lahman: Sean `Lahman' Baseball Database},
    author = {Michael Friendly and Chris Dalzell and Martin Monkman and Dennis Murphy},
    year = {2020},
    note = {R package version 8.0-0},
    url = {https://CRAN.R-project.org/package=Lahman},
  }

@article{gebru2021datasheets,
  title={Datasheets for datasets},
  author={Gebru, Timnit and Morgenstern, Jamie and Vecchione, Briana and Vaughan, Jennifer Wortman and Wallach, Hanna and Iii, Hal Daum{\'e} and Crawford, Kate},
  journal={Communications of the ACM},
  volume={64},
  number={12},
  pages={86--92},
  year={2021},
  publisher={ACM New York, NY, USA}
}

@Misc{rstanarm,
    title = {rstanarm: {Bayesian} applied regression modeling via {Stan}},
    author = {Ben Goodrich and Jonah Gabry and Imad Ali and Sam Brilleman},
    note = {R package version 2.21.3},
    year = {2022},
    url = {https://mc-stan.org/rstanarm/},
  }

@Manual{palmerpenguins,
    title = {palmerpenguins: Palmer Archipelago (Antarctica) penguin data},
    author = {Allison Marie Horst and Alison Presmanes Hill and Kristen B Gorman},
    year = {2020},
    note = {R package version 0.1.0},
    doi = {10.5281/zenodo.3960218},
    url = {https://allisonhorst.github.io/palmerpenguins/},
  }

@Article{rohan,
    title = {Welcome to the {tidyverse}},
    author = {Hadley Wickham and Mara Averick and Jennifer Bryan and Winston Chang and Lucy D'Agostino McGowan and Romain François and Garrett Grolemund and Alex Hayes and Lionel Henry and Jim Hester and Max Kuhn and Thomas Lin Pedersen and Evan Miller and Stephan Milton Bache and Kirill Müller and Jeroen Ooms and David Robinson and Dana Paige Seidel and Vitalie Spinu and Kohske Takahashi and Davis Vaughan and Claus Wilke and Kara Woo and Hiroaki Yutani},
    year = {2019},
    journal = {Journal of Open Source Software},
    volume = {4},
    number = {43},
    pages = {1686},
    doi = {10.21105/joss.01686},
  }

@Manual{,
  title = {readxl: Read Excel Files},
  author = {Hadley Wickham and Jennifer Bryan},
  year = {2023},
  note = {https://readxl.tidyverse.org, https://github.com/tidyverse/readxl},
}

@Article{,
    title = {Welcome to the {tidyverse}},
    author = {Hadley Wickham and Mara Averick and Jennifer Bryan and
      Winston Chang and Lucy D'Agostino McGowan and Romain François and
      Garrett Grolemund and Alex Hayes and Lionel Henry and Jim Hester
      and Max Kuhn and Thomas Lin Pedersen and Evan Miller and Stephan
      Milton Bache and Kirill Müller and Jeroen Ooms and David Robinson
      and Dana Paige Seidel and Vitalie Spinu and Kohske Takahashi and
      Davis Vaughan and Claus Wilke and Kara Woo and Hiroaki Yutani},
    year = {2019},
    journal = {Journal of Open Source Software},
    volume = {4},
    number = {43},
    pages = {1686},
    doi = {10.21105/joss.01686},
  }

@Article{,
    title = {Diagnostic Checking in Regression Relationships},
    author = {Achim Zeileis and Torsten Hothorn},
    journal = {R News},
    year = {2002},
    volume = {2},
    number = {3},
    pages = {7--10},
    url = {https://CRAN.R-project.org/doc/Rnews/},
  }

@Manual{sandwich,
  title = {sandwich: Robust Covariance Matrix Estimators},
  author = {Achim Zeileis},
  year = {2022},
  note = {R package version 3.0-1},
  url = {https://CRAN.R-project.org/package=sandwich},
}

@Article{,
    title = {Econometric Computing with {HC} and {HAC} Covariance
      Matrix Estimators},
    author = {Achim Zeileis},
    journal = {Journal of Statistical Software},
    year = {2004},
    volume = {11},
    number = {10},
    pages = {1--17},
    doi = {10.18637/jss.v011.i10},
  }

@Article{,
    author = {Hadley Wickham},
    title = {testthat: Get Started with Testing},
    journal = {The R Journal},
    year = {2011},
    volume = {3},
    pages = {5--10},
    url =
      {https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf},
  }

@Article{delondras2022impact,
  title = {The impact of mandatory waiting periods on abortion-related outcomes: a synthesis of legal and health evidence},
  author = {Fiona de Londras and Amanda Cleeve and Maria I. Rodriguez and Alana Farrell and Magdalena Furgalska and Antonella Lavelanet},
  journal = {BMC Public Health},
  volume = {22},
  pages = {1232},
  doi = {10.1186/s12889-022-13620-z},
  url = {https://bmcpublichealth.biomedcentral.com/articles/10.1186/s12889-022-13620-z},
}
